---
title: "Get Data"
description: >
  Learn how to get started with the primary functions.
output: html_vignette
vignette: >
  %\VignetteIndexEntry{Get Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(psrccensus)
library(dplyr)
```

## Get ACS data

Suppose you wish to tabulate ACS one-year 2019 data for estimates of total people by race and ethnicity, as provided in table B03002 by PSRC counties. You would use the following function call:

```{r acs examples, message=FALSE}
get_acs_recs(geography = 'county',
             table.names = 'B03002',
             years = 2019,
             acs.type = 'acs1')

```

### Default and custom arguments

By default, without specifying any counties, the jurisdictions returned will be King, Kitsap, Pierce, and Snohomish Counties. Use `?get_acs_recs()` for other default values implemented in this function.

To retrieve non-PSRC counties or a subset of the default counties, use the `counties` argument and provide a vector of counties (e.g. `counties = c("King", "Thurston")`). Do not use the `fips` argument as that is reserved for MSA or place geographies.

## Get Census data

The `get_decennial_recs()` to generate Decennial Census tables operates similarly to the `get_acs_recs()`. If you wanted to retrieve housing units and total population by MSA, you would call the following:

```{r census examples, message=FALSE}
get_decennial_recs(geography = 'msa',
                   table_codes = c("H001", "P001"),
                   years = c(2010),
                   fips = c('42660', "28420"))

```

Note: the table names are padded with 0s, so you call "H001" as opposed to "H1" as you would in Elmer. Only SF1 tables are currently implemented.
