---
title: "Getting Started"
author: "Data Science"
date: "`r format(Sys.time(), '%B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# psrccensus

The psrccensus package allows R users at PSRC to download, summarize, and visualize Census, ACS, and PUMS data on common PSRC Census geographies.

Some of the Census, ACS, and PUMS data can be found in our Elmer database: [see the internal wiki for more information](http://aws-linux/mediawiki/index.php/Elmer).

You would want to use this psrccensus R library as opposed to Elmer if: you are working in R, you need a table not in Elmer, and/or you would like a formatted table or map to be produced.

The three most important functions are:

-   [get_acs_recs(geography,state,counties,table.names.years, FIPS, acs.type)](https://psrc.github.io/psrccensus/reference/get_acs_recs.html "function for getting ACS tables")

-   [get_decennial_recs(geography, counties, table_codes,year, fips)](https://psrc.github.io/psrccensus/reference/get_decennial_recs.html "functions for getting Census tables")

-   get_pums_recs: to be coded later

We have one fancy function that let's create a map of an ACS variable by tract.

-   [create_tract_map( tract.tbl, tract.lyr, map.lat = 47.615, map.lon = -122.257, map.zoom = 8.5, wgs84 = 4326 )](https://psrc.github.io/psrccensus/reference/create_tract_map.html)

To use the library psrccensus, you first will need to get an api key.

### Set up your API key

The first time you run this code, you will need to set our Census API Key as an environment variable, if you haven't done that before. After that you can just get it. This is the website to get a key: <https://api.census.gov/data/key_signup.html>. Once you run Sys.setenv on the Census API Key you will only need to run Sys.getenv.

```{r keys, results='hide'}
library(psrccensus)
#Sys.setenv(CENSUS_API_KEY = 'PUT YOUR KEY HERE')
Sys.getenv("CENSUS_API_KEY")
```

Next you need to decide what tables you would like to download and summarize. This is the hardest part because you have find the correct table code, decide on geography, and which years.

## Finding the table ID

One helpful website for finding the tables on a topic from ACS and Census is: <https://censusreporter.org/>

Many of our frequently used Census/ACS/PUMS tables are nicely described in our wiki. Note that Census table names in Elmer are given a two character code such as H2.  In the R library, to be consistent with tidycensus the codes are padded with zeroes such as H002.

[2010 Census Table Names](http://aws-linux/mediawiki/index.php/2010_Decennial_Summary_Census_Datasets_in_Elmer)


[2019 ACS Table Names](http://aws-linux/mediawiki/index.php/Census-Year_2019_Summary_Datasets_in_Elmer)

You can also search the api variable lists from ACS and Census.

### 2019 ACS 1-year api Variable List

<https://api.census.gov/data/2019/acs/acs1/variables.html>

### 2015-2019 5 year- table list

<https://api.census.gov/data/2019/acs/acs5/variables.html>

### 2010 Census api variable list

<https://www.census.gov/data/developers/data-sets/decennial-census.html>

### Which variables are available in which year in ACS and Census?
Generally, ACS 1-year data are available down for geographies with populations of 65,000 or more. So you can easily get 1-year data for counties or the region, for example.  Once you want to go down to the tract-level, 5-year ACS data is more appropriate. Decennial Census data is available down to the block level.

# Get the data -----------------------------------------------------

The functions are documented on: <https://psrc.github.io/psrccensus/reference/index.html>

## Get ACS Data

Suppose you wish to tabulate ACS one year data 2019 data for estimates of total people by race and ethnicity, as provided in table B03002 by county. You would use the following function call.

```{r acs examples}
get_acs_recs(geography = 'county',
             table.names = c('B03002'),
             years=c(2019),
             acs.type = 'acs1')


```

## Get Census data

To generate Decennial Census tables for housing units and total population by MSA, you would call the following. Note: the table names are padded with 0s, so you call "H001" as opposed to "H1" as you would in Elmer.  Only SF1 tables are currently implemented.

```{r census examples}
get_decennial_recs(geography = 'msa',
                   table_codes = c("H001", "P001"),
                   year = 2010,
                   fips = c('42660', "28420"))

```

## Make a map of ACS data by tract

Let's say you want to create a map of the tracts in the region for one variable. You can use the function create_tract_map. Here's any example, mapping non-Hispanic Black or African American population alone by tract:

```{r tract_map example}
library(sf)
library(dplyr)
tract.big.tbl <- psrccensus::get_acs_recs(geography='tract',table.names=c('B03002'),years=c(2019))
tract.tbl<-tract.big.tbl %>% filter(label=='Estimate!!Total:!!Not Hispanic or Latino:!!Black or African American alone')
gdb.nm = paste0("MSSQL:server=","AWS-PROD-SQL\\Sockeye",
";database=","ElmerGeo",";trusted_connection=yes")
spn = 2285
tract_layer_name="dbo.tract2010_nowater"
tract.lyr <- st_read(gdb.nm, tract_layer_name, crs = spn)
create_tract_map(tract.tbl, tract.lyr, map.title='Black, non-Hispanic Population',
map.title.position='topleft', legend.title='Black, Non-Hispanic Population',
legend.subtitle='by Census Tract')

```
