---
title: "Getting Started"
description: >
  Learn what psrccensus is for and how to get set up for it
output: html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `psrccensus` package allows R users at PSRC to download, summarize, and visualize Census, ACS, and PUMS data on common PSRC Census geographies.

Some of the Census, ACS, and PUMS data can be found in our Elmer database: [see the internal wiki for more information](http://aws-linux/mediawiki/index.php/Elmer).

You would want to use `psrccensus` as opposed to Elmer if: you are working in R, you need a table not in Elmer, and/or you would like a formatted table or map to be produced.

## Primary Functions

The three most important functions are:

-   [`get_acs_recs()`](https://psrc.github.io/psrccensus/reference/get_acs_recs.html "function for getting ACS tables")

-   [`get_decennial_recs()`](https://psrc.github.io/psrccensus/reference/get_decennial_recs.html) "functions for getting Census tables")

-  [`get_psrc_pums()`](https://psrc.github.io/psrccensus/reference/get_psrc_pums.html) "function for getting Public Use Microsample Data")

We have one fancy function that lets us create a map of an ACS variable by tract.

-   [`create_tract_map()`](https://psrc.github.io/psrccensus/reference/create_tract_map.html)

To use `psrccensus`, you first will need to get an API key.

## Set up your API Key

A key is required to access the Census API. If you haven't done this before, you should set your Census API Key as an environment variable or store it in your .Renviron file. 

Obtain a Census API key here: <https://api.census.gov/data/key_signup.html>.

There are two methods to set your key:

1.  Once you run `Sys.setenv()` on the Census API Key you will only need to run `Sys.getenv()`.
    ```{r keys, results='hide'}
    library(psrccensus)
    #Sys.setenv(CENSUS_API_KEY = 'PUT YOUR KEY HERE')
    Sys.getenv("CENSUS_API_KEY")
    ```

2.  An alternative method is to install the key in your .Renviron File for repeated use. This is done via the `tidycensus` package. The benefit is that you would not have to call `Sys.getenv()` everytime you run your script or other scripts using `psrccensus` or `tidycensus`.

    ```{r keys2, eval=FALSE, results='hide'}
    # first time, run once in console. Then restart R
    tidycensus::census_api_key('your api key', install = TRUE)
    ```

Next you need to decide what tables you would like to download and summarize. This is the hardest part because you have to find the correct table code, decide on geography, and which years.

## Finding Tables and their Variables

One helpful website for finding the tables on a topic from the ACS and Census is: <https://censusreporter.org/>

Many of our frequently used Census/ACS/PUMS tables are nicely described in our wiki. Note that Census table names in Elmer are given a two character code such as "H2". In `psrccensus`, to be consistent with the `tidycensus` package, codes must be padded with zeroes so that "H2" becomes "H002".

-   [2010 Census Table Names](http://aws-linux/mediawiki/index.php/2010_Decennial_Summary_Census_Datasets_in_Elmer)
-   [2019 ACS Table Names](http://aws-linux/mediawiki/index.php/Census-Year_2019_Summary_Datasets_in_Elmer)

### Variable Lists

You can also search the API variable lists from ACS and Census.

-   **2019 ACS 1-year**: <https://api.census.gov/data/2019/acs/acs1/variables.html>

-   **2015-2019 ACS 5-year**: <https://api.census.gov/data/2019/acs/acs5/variables.html>

-   **2010 Census**: <https://www.census.gov/data/developers/data-sets/decennial-census.html>

### Which variables are available in which year in ACS and Census?

Generally, ACS 1-year data are available down for geographies with populations of 65,000 or more, so you can easily get 1-year data for counties or the region, for example. Once you want to go down to the tract-level, 5-year ACS data is more appropriate. Decennial Census data is available down to the block level.

Functions currently in `psrccensus` will only return data in **Washington State** at the **county, tract, MSA, and place** levels. To access other geographies, consider using `tidycensus` directly.

## Get Data

View the [Get Data](get-data.html) vignette (under Articles) on how to use the primary functions.

View the [Calculate PUMS Summaries](pums-1-basics.html) vignette (under Articles) on learn how to create aggregations of PUMS data for the region.
