---
title: "Getting Started"
author: "Data Science"
date: "`r format(Sys.time(), '%B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# psrccensus

The psrccensus package allows R users at PSRC to download, summarize, and visualize Census, ACS, and PUMS data on common PSRC Census geographies.

The three most important functions are:

-   [get_acs_recs(geography,state,counties,table.names.years, FIPS, acs.type)](https://psrc.github.io/psrccensus/reference/get_acs_recs.html "function for getting ACS tables")

-   [get_decennial_recs(geography, counties, table_codes,year, fips)](https://psrc.github.io/psrccensus/reference/get_decennial_recs.html "functions for getting Census tables")

-   get_pums_recs: to be coded later

To use the library psrccensus, you first will need to get an api key.

### Set up your API key

The first time you run this code, you will need to set our Census API Key as an environment variable, if you haven't done that before. After that you can just get it. This is the website to get a key: <https://api.census.gov/data/key_signup.html>. Once you run Sys.setenv on the Census API Key you will only need to run Sys.getenv.

```{r keys, results='hide'}
library(psrccensus)
#Sys.setenv(CENSUS_API_KEY = 'PUT YOUR KEY HERE')
Sys.getenv("CENSUS_API_KEY")
```

Next you need to decide what tables you would like to download and summarize. This is the hardest part because you have find the correct table code, decide on geography, and which years.

## Finding the table ID

One helpful website for finding the tables on a topic from ACS and Census is: <https://censusreporter.org/>

You can also search the api variable lists from ACS and Census.

### 2019 ACS 1-year api Variable List

<https://api.census.gov/data/2019/acs/acs1/variables.html>

### 2015-2019 5 year- table list

<https://api.census.gov/data/2019/acs/acs5/variables.html>

### 2010 Census api variable list

<https://www.census.gov/data/developers/data-sets/decennial-census.html>

# Get the data -----------------------------------------------------

The functions are documented on: <https://psrc.github.io/psrccensus/reference/index.html>

Suppose you wish to tabulate ACS one year data 2019 data for estimates of total people by race and ethnicity, as provided in table B03002 by county. You would use the following function call.

```{r acs examples}
get_acs_recs(geography = 'county',
             table.names = c('B03002'),
             years=c(2019),
             acs.type = 'acs1')


```

To generate Decennial Census tables for housing units and total population by MSA, you would call:

```{r census examples}
get_decennial_recs(geography = 'msa',
                   table_codes = c("H001", "P001"),
                   year = 2010,
                   fips = c('42660', "28420"))

```
